Budgeted Reinforcement Learning in Continuous State Space

### 2019-05-21 main.tex
Updates from Overleaf

\author{
Nicolas Carrara$^{1,}$\thanks{equal contribution}\and
Edouard Leurent$^{1,3,}$\footnotemark[1]\and
Romain Laroche$^{4}$\and
Tanguy Urvoy$^{2}$\and
Odalric Maillard$^{1}$\and
Olivier Pietquin$^{1,2,}$\thanks{now with Google Brain, Paris.}\\
\affiliations $^1$SequeL team, INRIA Lille -- Nord Europe, France\\
$^2$Orange Labs, Lannion, France\\
$^3$Renault Group, France\\
$^4$Microsoft Research , Montreal, Canada
\emails
\{nicolas.carrara, edouard.leurent, odalric.maillard, olivier.pietquin\}@inria.fr,
romain.laroche@microsoft.com, tanguy.urvoy@orange.com
}

@@

% @Nicolas/Edouard: faites ici la liste des sections qui sont pretes a etre relues.
% Section 1: il manque les contributions
% Section 2: OK. On pense éventuellement rajouter une hypothèse de contractance de Tau (pas sûr que ce soit le cas)
% Section 3: OK
% Section 4.1: OK dans un premier temps, mais on va peut-être modifier. En fait, on remplace pi_greedy(Q) par pi_hull(Q), et on ne peut faire cela que lorsque Q est concave. Deux possibilité: soit on dit juste que Tau_greedy est remplacé par Tau_hull, qui admet également Q* comme point fixe (Tau_greedy(Q*) = Tau_hull(Q*) puisque Q* concave d'après le lemme). Soit on est intéressé par ce qui se passe le long de la trajectoire Q_k (par exemple si on suppose Tau_greedy contractant), et pour conserver cette propriété il faudrait alors montrer que Tau conserve la concavité et que donc on a bien Tau_hull = Tau_greedy tout le long de la trajectoire. Il faut aussi reprendre les notations des graphiques.
% Section 4.2 et 4.3: KO. Pas encore retravaillé, mais ça ne devrait pas trop bougé
% Section 5: OK pour environnements et results, KO pour discussion
% Section 6: KO
% Section 7: KO

@@

Reinforcement Learning (RL) is a general framework for decision-making under uncertainty. It frames the learning objective as the optimal control of a Markov Decision Process  $(\cS, \cA, P, R, \gamma)$ with continuous state space $\cS$, discrete actions $\cA$, and unknown dynamics $P\in \cM(\cS)^{\cS \times \cA}$ and rewards $R\in\Real^{\cS \times \cA}$, where $\mathcal{M}(\mathcal{X})$ denotes the set of probability measures over a measurable space $\mathcal{X}$. Formally, we seek a policy $\pi\in\cM(A)^\cS$ that maximises in expectation the $\gamma$-discounted return of rewards $G_r^\pi = \sum_{t=0}^\infty \gamma^t R(s_t, a_t)$.

@@

Reinforcement Learning also assumes that the performance can be described solely by a reward function $R$. Conversely, real problems typically involve many aspects, some of which can be contradictory \citep{ChunmingLiu2014}. For instance, a self-driving car needs to balance between progressing quickly on the road and avoiding collisions. When aggregating several objectives in a single scalar signal, as often in Multi-Objectives RL~\citep{Roijers2013ASO}, no control is given over their relative ratios, as high rewards can compensate high penalties. For instance, if a weighted sum is used to balance velocity $v$ and accidents $a$, then for any given choice of weights $\omega$ the optimality equation $\omega_v\expectedvalue[\sum\gamma^t v_t] + \omega_a\expectedvalue[\sum\gamma^t a_t] = G_r^* = \max_\pi G^\pi_r$ is the equation of a line in $(\expectedvalue[\sum\gamma^t v_t], \expectedvalue[\sum\gamma^t a_t])$, and the automotive company cannot control where its optimal policy $\pi^*$ lies on that line.

Both of these concerns can be addressed in the \emph{Constrained Markov Decision Process} (CMDP) setting \citep{BEUTLER1985236,Altman95constrainedmarkov}. In this multi-objective formulation, task completion and safety are considered separately. We equip the MDP with a cost signal $C \in \Real^{\cS\times \cA}$ and a cost budget $\beta\in\Real$. Similarly to $G_r$, we define the return of costs $G_c^\pi = \sum_{t=0}^\infty \gamma^t C(s_t, a_t)$ and the new cost-constrained objective:
\begin{equation}
\label{eq:cmdp}
\begin{array}{lcr}
 \max_{\pi\in\cM(\cA)^\cS} \expectedvalue[G_r^\pi | s_0=s, \pi] & \text{ s.t. } & \expectedvalue[G_c^\pi | s_0=s, \pi] \leq \beta
\end{array}
\end{equation}

@@

To address this concern, the \emph{Budgeted Markov Decision Process} (BMDP) was introduced in \citep{Boutilier_Lu:uai16} as an extension of CMDPs to enable the online control over the budget $\beta$ within an interval $\cB \subset \Real$ of admissible budgets. Instead of fixing the budget prior to training, the objective is now to find a generic optimal policy $\pi^*$ takes $\beta$ as input so as to solve the corresponding CMDP \eqref{eq:cmdp} for all $\beta\in\cB$. This gives the system designer the ability to move the optimal policy $\pi^*$ in real-time along the Pareto-optimal curve of the different reward-cost trade-offs.

Our first contribution is to re-frame the original BMDP formulation in the context of continuous states and infinite discounted horizon. We then propose a novel Budgeted Bellman Optimality Operator and prove the optimal value function to be a fixed point of this operator. Second, we use this operator in \BFTQ, a Reinforcement Learning algorithm for solving BMDPs online by interaction with an environment, through function approximation and a tailored exploration procedure. Third, we scale this algorithms to large problems by providing an efficient implementation of the Budgeted Bellman Optimality Operator based on convex computing, and by leveraging tools from Deep Reinforcement Learning such as Deep Neural Networks and synchronous parallel computing. Finally, we validate our approach in two environments that display a clear trade-off between rewards and costs: a spoken dialogue system and a problem of behaviour planning for autonomous driving.

@@

For ease of notation, we consider the \emph{augmented} state and action spaces $\ocS = \cS\times \cB$ and $\ocA= \cA\times \cB$, and equip them with the augmented dynamics $\ov{P}\in \cM(\ocS)^{\ocS \times \ocA}$ defined as:
\begin{equation*}
    \ov{P}\left(\os' \condbar \os, \oa\right) = \ov{P}\left((s',\beta') \condbar (s,\beta), (a, \beta_a)\right) = P(s'|s, a)\delta(\beta' - \beta_a)
\end{equation*}
where $\delta$ is the Dirac distribution.

@@

We stack the rewards and cost functions in a single \emph{vectorial} signal $\ov{R} \in (\Real^2)^{{\ocS \times \ocA}}$.
Given an augmented transition $(\os, \oa) =((s,\beta), (a, \beta_a))$, we define:
\begin{equation}
     \ov{R}(\os, \oa) =  \begin{bmatrix}
     R(s, a)\\
     C(s, a)
     \end{bmatrix}\in\Real^2
\end{equation}

As earlier, the return $G^\pi = (G_r^\pi, G_c^\pi)$ of a budgeted policy $\pi\in\Pi$ refers to:
% \begin{equation*}
    $G^\pi = \sum_{t=0}^\infty \gamma^t \ov{R}(\os_t, \oa_t)$.
% \end{equation*}

We can introduce the value function $V^\pi$ and action-value function $Q^\pi$ of a budgeted policy $\pi\in\Pi$ as:
% \begin{align}
% \label{eq:value-function}
%     V^\pi(\os) &= \expectedvalue\left[ G^\pi \condbar \ov{s_0} = \os\right] \\
%     Q^\pi(\os, \oa) &= \expectedvalue\left[ G^\pi \condbar \ov{s_0} = \os, \ov{a_0} = \oa\right]
% \end{align}
\begin{equation}
    \label{eq:value-function}
V^\pi(\os) = \expectedvalue\left[ G^\pi \condbar \ov{s_0} = \os\right] \qquad Q^\pi(\os, \oa) = \expectedvalue\left[ G^\pi \condbar \ov{s_0} = \os, \ov{a_0} = \oa\right]
\end{equation}

@@

\begin{proposition}[Budgeted Bellman Expectation]
\label{prop:bellman-expectation}
The value functions $V^\pi, Q^\pi$ of a budgeted policy $\pi$ verify:
\begin{align}
    V^\pi(\os) &= \sum_{\oa\in\ocA}\pi(\oa | \os) Q^\pi(\os, \oa)\\
    Q^\pi(\os, \oa) &= \ov{R}(\os, \oa) + \gamma\sum_{\os'\in\ocS}\ov{P}\left(\os' \condbar \os, \oa\right) V^\pi(\os') \label{eq:bellman_expectation_Q}
\end{align}
\end{proposition}

\begin{definition}[Budgeted Optimality]
We now come to the definition of budgeted optimality. We want an optimal budgeted policy to: (i)~respect the cost budget $\beta$, (ii)~maximise the rewards $R$, (iii)~in case of tie, minimise the costs $C$. To that end, we define the following sets for all $\os\in\ocS$:
\begin{itemize}
    \item Admissible policies: $\Pi_a(\os) = \{\pi\in\Pi: V_c^\pi(\os) \leq \beta\}$ where $\os = (s, \beta)$.
    \item Best admissible policies: $\Pi_r(\os) = \argmax_{\pi\in\Pi_a(\os)}  V_r^\pi(\os)$
    \item Optimal policies: $\Pi^*(\os) = \argmin_{\pi\in\Pi_r(\os)}  V_c^\pi(\os)$
\end{itemize}
\end{definition}

@@

Let $\pi^*$ be one such optimal budgeted policy, we finally define the optimal budgeted value functions $V^*$, $Q^*$. For all $\os, \oa \in \ocS\times\ocA$:
\begin{equation*}
    V^*(\os) = V^{\pi^*}(\os), \qquad Q^*(\os, \oa) =  Q^{\pi^*}(\os, \oa),
\end{equation*}

@@

\begin{theorem}[Budgeted Bellman Optimality]
\label{thm:bellman-optimality}
The optimal action-value of a budgeted policy verifies:
\begin{equation}
\label{eq:bellman-optimality}
    Q^{*}(\os, \oa) = \cT Q^{*}(\os, \oa) \eqdef \ov{R}(\os, \oa) + \gamma \sum_{\os'\in\ocS} \ov{P}(\ov{s'} | \os, \oa)\sum_{\ov{a'}\in A} \pi_\text{greedy}(\ov{a'}|\ov{s'}; Q^*) Q^{*}(\ov{s'}, \ov{a'})
\end{equation}
where the greedy policy $\pi_\text{greedy}$ is defined by: $\forall \os=(s,\beta)\in \ocS, \oa\in 
\ocA, \forall Q\in(\Real^2)^{\ocA\times\ocS},$
\begin{subequations}
\label{eq:pi_greedy}
\begin{equation}
    \pi_\text{greedy}(\oa|\os; Q) \in \argmin_{\rho\in\Pi_r^Q} \expectedvalueover{\oa\sim\rho}Q_c(\os, \oa) \label{eq:pi_greedy_cost}
\end{equation}
\begin{align}
    \text{where }\quad\Pi_r^Q = &\argmax_{\rho\in\cM(\ocA)} \expectedvalueover{\oa\sim\rho} Q_r(\os, \oa) \label{eq:pi_greedy_reward}\\
    & \text{ s.t. }  \expectedvalueover{\oa\sim\rho} Q_c(\os, \oa) \leq \beta \label{eq:pi_greedy_constraint}
\end{align}
\end{subequations}
\end{theorem}

@@

\begin{remark}[Notations]
Because of the budget dynamics $\beta' = \beta_a$, we can see in \eqref{eq:bellman_expectation_Q} that for all $\os=(s, \beta)$ and $\oa=(a, \beta_a)$, $Q^\pi(\os, \oa)$ only depends on $s, a, \beta_a$ and not on $\beta$. We will slightly abuse notations and denote in the following $Q^\pi(s, a, \beta_a) \eqdef Q^\pi(\os, \oa)$.
\end{remark}

@@

When the BMDP is unknown, we need to adapt Algorithm~\ref{algo:bvi} to work with a batch of samples $\cD=\{(s_i,\beta_i,a_i,R_i,C_i,s_i')\}_{i\in [0,N]}$ collected by interaction with the environment. Applying $\cT$ in \eqref{eq:bellman-optimality} requires computing an expectation $\expectedvalueover{\os'\sim \ov{P}}$ over next states $\os'$, which we can replace by an empirical average $\sum_{s'\in\cD}$.
We introduce in Algorithm~\ref{algo:bftq} the \emph{Budgeted-Fitted-Q} (BFTQ) algorithm, an extension of the \emph{Fitted-Q} (\FTQ) algorithm \citep{Ernst2005,Riedmiller2005} adapted to solve unknown BMDPs. 
Because we work on samples of a continuous state space $\cS$, we need to employ function-approximation in order to generalise to nearby states. Precisely, given a parametrized model $Q_\theta$, we seek to minimise a regression loss $L(Q_\theta, Q_\text{target};\cD) = \sum_{\cD} ||Q_\theta(\os, \oa) - Q_\text{target}(\os, \oa, \ov{R}, \os')||_2^2$.
Any model can be used, such as linear models, regression trees, or neural networks.

\begin{minipage}[t]{0.5\textwidth}
\vspace{0pt}  
\begin{algorithm}[H]
\label{algo:bvi}
\DontPrintSemicolon
\KwData{$P, R, C$}
\KwResult{$Q^*$}
$Q_{0} \leftarrow 0$\;
\Repeat{convergence}{
    $Q_{k+1} \leftarrow \cT Q_k$\;
}
\caption{Budgeted Value Iteration}
\end{algorithm}
\end{minipage}%
\begin{minipage}[t]{0.5\textwidth}
\vspace{0pt}
\begin{algorithm}[H]
\label{algo:bftq}
\DontPrintSemicolon
\KwData{$\cD$}
\KwResult{$Q^*$}
$Q_{0} \leftarrow 0$\;
\Repeat{convergence}{
    $Q_{k+1} \leftarrow \argmin_\theta \cL(Q_\theta, \cT Q_k; \cD)$\;
}
\caption{Budgeted Fitted-Q}
\end{algorithm}
\end{minipage}

@@

As stated in Remark~\ref{rmk:greedy}, computing the greedy policy $\pi_\text{greedy}$ in \eqref{eq:bellman-optimality} is not trivial since it requires solving the nested non-linear constrained optimisation program \eqref{eq:pi_greedy}. However, it can be solved efficiently by exploiting the \emph{structure} of the optimal value function $V^*$. % In the following, we say that a vector-valued function is concave and non-decreasing with respect to a variable if all of its components are.

%\begin{lemma}[Concavity and monotony of $Q^*$]
%\label{lemma:concavity}
%$Q^*$ is non-decreasing concave with respect to $\beta_a$. 
%\end{lemma}
%\begin{lemma}[Conservation of concavity and monotony by $\cT$]
%\label{lemma:tau_concavity}
%For any $Q\in(\Real^2)^{\ocS\ocA}$ non-decreasing concave with respect to $\beta_a$, $\cT Q$ is also non-decreasing concave with respect to $\beta_a$.
%\end{lemma}

%These two Lemmas can be used to derive an efficient computation method for $\pi_\text{greedy}$.

%\begin{lemma}[Hull policy equality]
%\label{lemma:pi_hull}
%For any $Q\in(\Real^2)^{\ocS\ocA}$ non-decreasing concave with respect to $\beta_a$, $\pi_\text{greedy}$ in \eqref{eq:pi_greedy} can be computed by running Algorithm~\ref{algo:pi_hull}.
%\begin{equation*}
%    \pi_\text{greedy}(\oa|\os; Q) = \pi_\text{hull}(\oa|\os; Q)
%\end{equation*}
%\end{lemma}

@@

\begin{algorithm}
\DontPrintSemicolon
\KwData{$s$, $Q$}
$\cH \leftarrow \texttt{convex\_hull}(Q(s,\cA, \cB))$\;
denote $q^i = Q(s,a^i,\beta_a^i)\in\cH$ its $i^{\text{th}}$ point in clockwise order\;
find two successive points $((q_c^0, q_r^0), (q_c^1, q_r^1))$ of $\cH$ such that $q_c^0 \leq \beta < q_c^1$\;
\lIf{$q_r^0 = q_r^1$}{
\Return $\delta(a^0, \beta_a^0)$
}\Else{
$p \leftarrow (\beta - q_c^0) / (q_c^1 - q_c^0)$\;
\Return the mixture $(1-p)\delta(a^0, \beta_a^0) + p\delta(a^1, \beta_a^1)$\;
}
\caption{Convex hull policy $\pi_\text{hull}(\oa|\os; Q)$}
\label{algo:pi_hull}
\end{algorithm}

\begin{minipage}[t]{0.5\textwidth}
\vspace{0pt}
\begin{figure}[H]
    \centering
    \includegraphics[width=\linewidth]{source/img/qrqc}
    \caption{Computation of $\pi_\text{hull}$ for a concave non-decreasing Q-function $Q$. The dotted tangent on the right defines the greedy policy when the budget $\beta$ lies between $Q_c(s,a^0,\beta_a^0)$ and $Q_c(s,a^1,\beta_a^1)$, as mixtures of the two points $(0, 1)$ represented as green crosses}
    \label{fig:hull}
\end{figure}
\end{minipage}
\begin{minipage}[t]{0.5\textwidth}
\begin{figure}[H]
    \centering
    \input{source/architecture.tex}
    \caption{Neural Network for $Q$-functions approximation when the state dimension is 2 and there are 2 actions.}
    \label{fig:architecture}
\end{figure}
\end{minipage}

@@

Neural networks are well suited to model Q-functions in Reinforcement Learning algorithms \citep{Riedmiller2005,Mnih2015}.  We approximate $Q = (Q_r, Q_c)$ using one single neural network. Thus, the ywo components are jointly optimised which accelerates convergence and fosters learning of useful shared representations. Moreover, as in \citep{Mnih2015} we are dealing with a finite (categorical) action space $\cA$, instead of including the action in the input we add the output of the $Q$-function for each action to the last layer. Again, it provides a faster convergence toward useful shared representations and it only requires one forward pass to evaluate all action values. Finally, beside the state $s$ there is one more input to a budgeted $Q$-function:~the budget $\beta_a$. This budget is a scalar value whereas the state $s$ is a vector of potentially large size. To avoid a weak influence of $\beta$ compared to $s$ in the prediction, we include an additional encoder for the budget, whose width and depth may depend on the application. A straightforward choice is a single layer with the same width as the state. The overall architecture is shown in Fig.~\ref{fig:architecture}.

>>>>

\author{
Nicolas Carrara$^{1,}$\thanks{equal contribution}\and
Edouard Leurent$^{1,3,}$\footnotemark[1]\and
Romain Laroche$^{4}$\and
Tanguy Urvoy$^{2}$\and
Odalric Maillard$^{1}$\and
Olivier Pietquin$^{1,2,}$\thanks{now with Google Research - Brain Team, Paris, France.}\\
\affiliations $^1$SequeL team, INRIA Lille -- Nord Europe, France\\
$^2$Orange Labs, Lannion, France\\
$^3$Renault Group, France\\
$^4$Microsoft Research , Montreal, Canada
\emails
\{nicolas.carrara, edouard.leurent, odalric.maillard, olivier.pietquin\}@inria.fr,
romain.laroche@microsoft.com, tanguy.urvoy@orange.com
}

@@

% @Nicolas/Edouard: faites ici la liste des sections qui sont pretes a etre relues.
% Section 1: il manque les contributions
% Section 2: OK. On pense éventuellement rajouter une hypothèse de contractance de Tau (pas sûr que ce soit le cas)
% Section 3: OK
% Section 4: OK. Il faut aussi reprendre les notations des graphiques.
% Section 5: OK
% Section 6: OK
% Section 7: OK

@@

Reinforcement Learning (RL) is a general framework for decision-making under uncertainty. It frames the learning objective as the optimal control of a Markov Decision Process  $(\cS, \cA, P, R_r, \gamma)$ with continuous state space $\cS$, discrete actions $\cA$, and unknown dynamics $P\in \cM(\cS)^{\cS \times \cA}$ and rewards $R_r\in\Real^{\cS \times \cA}$, where $\mathcal{M}(\mathcal{X})$ denotes the set of probability measures over a measurable space $\mathcal{X}$. Formally, we seek a policy $\pi\in\cM(A)^\cS$ that maximises in expectation the $\gamma$-discounted return of rewards $G_r^\pi = \sum_{t=0}^\infty \gamma^t R_r(s_t, a_t)$.

@@

Reinforcement Learning also assumes that the performance can be described solely by a reward function $R_r$. Conversely, real problems typically involve many aspects, some of which can be contradictory \citep{ChunmingLiu2014}. For instance, a self-driving car needs to balance between progressing quickly on the road and avoiding collisions. When aggregating several objectives in a single scalar signal, as often in Multi-Objectives RL~\citep{Roijers2013ASO}, no control is given over their relative ratios, as high rewards can compensate high penalties. For instance, if a weighted sum is used to balance velocity $v$ and accidents $a$, then for any given choice of weights $\omega$ the optimality equation $\omega_v\expectedvalue[\sum\gamma^t v_t] + \omega_a\expectedvalue[\sum\gamma^t a_t] = G_r^* = \max_\pi G^\pi_r$ is the equation of a line in $(\expectedvalue[\sum\gamma^t v_t], \expectedvalue[\sum\gamma^t a_t])$, and the automotive company cannot control where its optimal policy $\pi^*$ lies on that line.

Both of these concerns can be addressed in the \emph{Constrained Markov Decision Process} (CMDP) setting \citep{BEUTLER1985236,Altman95constrainedmarkov}. In this multi-objective formulation, task completion and safety are considered separately. We equip the MDP with a cost signal $R_c \in \Real^{\cS\times \cA}$ and a cost budget $\beta\in\Real$. Similarly to $G_r$, we define the return of costs $G_c^\pi = \sum_{t=0}^\infty \gamma^t R_c(s_t, a_t)$ and the new cost-constrained objective:
\begin{equation}
\label{eq:cmdp}
\begin{array}{lcr}
 \max_{\pi\in\cM(\cA)^\cS} \expectedvalue[G_r^\pi | s_0=s, \pi] & \text{ s.t. } & \expectedvalue[G_c^\pi | s_0=s, \pi] \leq \beta
\end{array}
\end{equation}

@@

To address this concern, the \emph{Budgeted Markov Decision Process} (BMDP) was introduced in \citep{Boutilier_Lu:uai16} as an extension of CMDPs to enable the online control over the budget $\beta$ within an interval $\cB \subset \Real$ of admissible budgets. Instead of fixing the budget prior to training, the objective is now to find a generic optimal policy $\pi^*$ that takes $\beta$ as input so as to solve the corresponding CMDP \eqref{eq:cmdp} for all $\beta\in\cB$. This gives the system designer the ability to move the optimal policy $\pi^*$ in real-time along the Pareto-optimal curve of the different reward-cost trade-offs.

Our first contribution is to re-frame the original BMDP formulation in the context of continuous states and infinite discounted horizon. We then propose a novel Budgeted Bellman Optimality Operator and prove the optimal value function to be a fixed point of this operator. Second, we use this operator in a Reinforcement Learning algorithm for solving BMDPs online by interaction with an environment, through function approximation and a tailored exploration procedure. Third, we scale this algorithms to large problems by providing an efficient implementation of the Budgeted Bellman Optimality Operator based on convex computing, and by leveraging tools from Deep Reinforcement Learning such as Deep Neural Networks and synchronous parallel computing. Finally, we validate our approach in two environments that display a clear trade-off between rewards and costs: a spoken dialogue system and a problem of behaviour planning for autonomous driving.

The proofs of our main results are provided in Supplementary Material \ref{sec:proofs}.

@@

For ease of notation, we consider the \emph{augmented} state and action spaces $\ocS = \cS\times \cB$ and $\ocA= \cA\times \cB$, and equip them with the augmented dynamics $\ov{P}\in \cM(\ocS)^{\ocS \times \ocA}$ defined as:
\begin{equation*}
    \ov{P}\left(\os' \condbar \os, \oa\right) = \ov{P}\left((s',\beta') \condbar (s,\beta), (a, \beta_a)\right) \eqdef P(s'|s, a)\delta(\beta' - \beta_a),
\end{equation*}
where $\delta$ is the Dirac distribution.

@@

We stack the rewards and cost functions in a single \emph{vectorial} signal $R \in (\Real^2)^{{\ocS \times \ocA}}$.
Given an augmented transition $(\os, \oa) =((s,\beta), (a, \beta_a))$, we define:
\begin{equation}
     R(\os, \oa) \eqdef  \begin{bmatrix}
     R_r(s, a)\\
     R_c(s, a)
     \end{bmatrix}\in\Real^2.
\end{equation}

As earlier, the return $G^\pi = (G_r^\pi, G_c^\pi)$ of a budgeted policy $\pi\in\Pi$ refers to:
    $G^\pi \eqdef \sum_{t=0}^\infty \gamma^t R(\os_t, \oa_t)$.

We can introduce the value function $V^\pi$ and action-value function $Q^\pi$ of a budgeted policy $\pi\in\Pi$ as:
\begin{equation}
    \label{eq:value-function}
V^\pi(\os) \eqdef \expectedvalue\left[ G^\pi \condbar \ov{s_0} = \os\right] \qquad Q^\pi(\os, \oa) \eqdef \expectedvalue\left[ G^\pi \condbar \ov{s_0} = \os, \ov{a_0} = \oa\right].
\end{equation}

@@

\begin{proposition}[Budgeted Bellman Expectation]
\label{prop:bellman-expectation}
The value functions $V^\pi, Q^\pi$ of a budgeted policy $\pi$ verify:
\begin{align}
    V^\pi(\os) &= \sum_{\oa\in\ocA}\pi(\oa | \os) Q^\pi(\os, \oa),\\
    Q^\pi(\os, \oa) &= R(\os, \oa) + \gamma\sum_{\os'\in\ocS}\ov{P}\left(\os' \condbar \os, \oa\right) V^\pi(\os') \label{eq:bellman_expectation_Q}.
\end{align}
\end{proposition}

\begin{definition}[Budgeted Optimality]
We now come to the definition of budgeted optimality. We want an optimal budgeted policy to: (i)~respect the cost budget $\beta$, (ii)~maximise the rewards $R_r$, (iii)~in case of tie, minimise the costs $R_c$. To that end, we define the following sets for all $\os\in\ocS$:
\begin{itemize}
    \item Admissible policies: $\Pi_a(\os) \eqdef \{\pi\in\Pi: V_c^\pi(\os) \leq \beta\}$ where $\os = (s, \beta)$.
    \item Best admissible policies: $\Pi_r(\os) \eqdef \argmax_{\pi\in\Pi_a(\os)}  V_r^\pi(\os)$.
    \item Optimal policies: $\Pi^*(\os) \eqdef \argmin_{\pi\in\Pi_r(\os)}  V_c^\pi(\os)$.
\end{itemize}
\end{definition}

@@

Let $\pi^*$ be one such optimal budgeted policy, we finally define the optimal budgeted value functions $V^*$, $Q^*$. For all $\os, \oa \in \ocS\times\ocA$:
\begin{equation*}
    V^*(\os) \eqdef V^{\pi^*}(\os), \qquad Q^*(\os, \oa) \eqdef  Q^{\pi^*}(\os, \oa),
\end{equation*}

@@

\begin{theorem}[Budgeted Bellman Optimality]
\label{thm:bellman-optimality}
The optimal action-value of a budgeted policy verifies:
\begin{equation}
\label{eq:bellman-optimality}
    Q^{*}(\os, \oa) = \cT Q^{*}(\os, \oa) \eqdef R(\os, \oa) + \gamma \sum_{\os'\in\ocS} \ov{P}(\ov{s'} | \os, \oa)\sum_{\ov{a'}\in A} \pi_\text{greedy}(\ov{a'}|\ov{s'}; Q^*) Q^{*}(\ov{s'}, \ov{a'}),
\end{equation}
where the greedy policy $\pi_\text{greedy}$ is defined by: $\forall \os=(s,\beta)\in \ocS, \oa\in 
\ocA, \forall Q\in(\Real^2)^{\ocA\times\ocS},$
\begin{subequations}
\label{eq:pi_greedy}
\begin{equation}
    \pi_\text{greedy}(\oa|\os; Q) \in \argmin_{\rho\in\Pi_r^Q} \expectedvalueover{\oa\sim\rho}Q_c(\os, \oa), \label{eq:pi_greedy_cost}
\end{equation}
\begin{align}
    \text{where }\quad\Pi_r^Q \eqdef &\argmax_{\rho\in\cM(\ocA)} \expectedvalueover{\oa\sim\rho} Q_r(\os, \oa) \label{eq:pi_greedy_reward}\\
    & \text{ s.t. }  \expectedvalueover{\oa\sim\rho} Q_c(\os, \oa) \leq \beta. \label{eq:pi_greedy_constraint}
\end{align}
\end{subequations}
\end{theorem}

@@

\begin{remark}[Notations]
Because of the budget dynamics $\beta' = \beta_a$, we can see in \eqref{eq:bellman_expectation_Q} that for all $\os=(s, \beta)$ and $\oa=(a, \beta_a)$, $Q^\pi(\os, \oa)$ only depends on $s, a, \beta_a$ and not on $\beta$. We will slightly abuse notations and sometimes denote in the following $Q^\pi(s, a, \beta_a) \eqdef Q^\pi(\os, \oa)$.
\label{remark:notations}
\end{remark}

@@

When the BMDP is unknown, we need to adapt Algorithm~\ref{algo:bvi} to work with a batch of samples $\cD=\{(\os_i=(s_i,\beta_i),\oa_i=(a_i,\beta_{a_i}),R_i=({R_r}_i,{R_c}_i),\os_i'=(s_i',\beta_{a_i})\}_{i\in [0,N]}$ collected by interaction with the environment. Following \cref{remark:notations}, we rewrite this batch as  $\cD=\{s_i,a_i,\beta_{a_i},r_i,c_i,s_i'\}_{i\in [0,N]}$. Applying $\cT$ in \eqref{eq:bellman-optimality} would require computing an expectation $\expectedvalue_{\os'\sim \ov{P}}$ over next states $\os'$ and hence an access to the model $\ov{P}$. We instead use $\hat{\cT}$ in which this expectation is replaced by a sample average:
\begin{equation*}
    \hat{\cT} Q(\os, \oa, R, \os') \eqdef R + \gamma \sum_{\ov{a'}\in A} \pi_\text{greedy}(\ov{a'}|\ov{s'}; Q) Q(\ov{s'}, \ov{a'}).
\end{equation*}
We introduce in Algorithm~\ref{algo:bftq} the \emph{Budgeted-Fitted-Q} (BFTQ) algorithm, an extension of the \emph{Fitted-Q} (\FTQ) algorithm \citep{Ernst2005,Riedmiller2005} adapted to solve unknown BMDPs. Because we work with  continuous state space $\cS$ and budget space $\cB$, we need to employ function-approximation in order to generalise to nearby states and budgets. Precisely, given a parametrized model $Q_\theta$, we seek to minimise a regression loss $\cL(Q_\theta, Q_\text{target};\cD) = \sum_{\cD} ||Q_\theta(\os, \oa) - Q_\text{target}(\os, \oa, R, \os')||_2^2$.
Any model can be used, such as linear models, regression trees, or neural networks.

\begin{minipage}[t]{0.47\textwidth}
\vspace{0pt}  
\begin{algorithm}[H]
\label{algo:bvi}
\DontPrintSemicolon
\KwData{$P, R_r, R_c$}
\KwResult{$Q^*$}
$Q_{0} \leftarrow 0$\;
\Repeat{convergence}{
    $Q_{k+1} \leftarrow \cT Q_k$\;
}
\caption{Budgeted Value Iteration}
\end{algorithm}
\end{minipage}%
\hfill
\begin{minipage}[t]{0.47\textwidth}
\vspace{0pt}
\begin{algorithm}[H]
\label{algo:bftq}
\DontPrintSemicolon
\KwData{$\cD$}
\KwResult{$Q^*$}
$Q_{\theta_0} \leftarrow 0$\;
\Repeat{convergence}{
    $\theta_{k+1} \leftarrow \argmin_\theta \cL(Q_\theta, \hat{\cT} Q_{\theta_{k}}; \cD)$\;
}
\caption{Budgeted Fitted-Q}
\end{algorithm}
\end{minipage}

@@

As stated in Remark~\ref{rmk:greedy}, computing the greedy policy $\pi_\text{greedy}$ in \eqref{eq:bellman-optimality} is not trivial since it requires solving the nested non-linear constrained optimisation program \eqref{eq:pi_greedy}. However, it can be solved efficiently by exploiting the \emph{structure} of the set of solutions with respect to $\beta$, which is concave and increasing.

@@

\begin{algorithm}
\DontPrintSemicolon
\KwData{$s$, $Q$}
$Q^+\leftarrow \{Q_c > \min \{Q_c(s,\oa) \text{ s.t. }\oa\in\argmax_{\oa} Q_r(s,\oa)\} \}$\tcp*[f]{dominated points}\;
$\cF \leftarrow \text{top frontier of }\texttt{convex\_hull}(Q(s,\ocA) \setminus Q^+)$\tcp*[f]{candidate mixtures}\;
$\cF_Q \leftarrow \cF\cap Q(s,\ocA)$\;
\For{points $q = Q(s,\oa)\in\cF_Q$ in clockwise order}{
\uIf{find two successive points $((q_c^1, q_r^1), (q_c^2, q_r^2))$ of $\cF_Q$ such that $q_c^1 \leq \beta < q_c^2$}{
$p \leftarrow (\beta - q_c^1) / (q_c^2 - q_c^1)$\;
\Return the mixture $(1-p)\delta(\oa-\oa^1) + p\delta(\oa-\oa^2)$\;
}}
\lElse{\Return $\delta(\oa - \argmax_{\oa} Q_r(s,\oa))$\tcp*[f]{Budget $\beta$ always respected}}
\caption{Convex hull policy $\pi_\text{hull}(\oa|\os; Q)$}
\label{algo:pi_hull}
\end{algorithm}

The computation of $\pi_\text{hull}$ in \cref{algo:pi_hull} is illustrated in \cref{fig:hull}.

\begin{minipage}[t]{0.5\textwidth}
\vspace{0pt}
\begin{figure}[H]
    \centering
    \includegraphics[width=\linewidth]{source/img/pi.pdf}
    \caption{Representation of $\pi_\text{hull}$. When the budget lies between $Q(\os,\oa_1)$ and $Q(\os,\oa_2)$, two points of the top frontier of the convex hull, then the policy is a mixture of these two points.}
    \label{fig:hull}
\end{figure}
\end{minipage}
\begin{minipage}[t]{0.5\textwidth}
\begin{figure}[H]
    \centering
    \input{source/architecture.tex}
    \caption{Neural Network for $Q$-functions approximation when $\cS=\Real^2$ and $|\cA| = 2$.}
    \label{fig:architecture}
\end{figure}
\end{minipage}

@@

Neural networks are well suited to model Q-functions in Reinforcement Learning algorithms \citep{Riedmiller2005,Mnih2015}.  We approximate $Q = (Q_r, Q_c)$ using one single neural network. Thus, the two components are jointly optimised which accelerates convergence and fosters learning of useful shared representations. Moreover, as in \citep{Mnih2015} we are dealing with a finite (categorical) action space $\cA$, instead of including the action in the input we add the output of the $Q$-function for each action to the last layer. Again, it provides a faster convergence toward useful shared representations and it only requires one forward pass to evaluate all action values. Finally, beside the state $s$ there is one more input to a budgeted $Q$-function:~the budget $\beta_a$. This budget is a scalar value whereas the state $s$ is a vector of potentially large size. To avoid a weak influence of $\beta$ compared to $s$ in the prediction, we include an additional encoder for the budget, whose width and depth may depend on the application. A straightforward choice is a single layer with the same width as the state. The overall architecture is shown in Fig.~\ref{fig:architecture}.
